---
title: "Lab 3"
author: "Eric Gleiter; Brandon Xicon"
date: "2/19/2020"
output: html_document
---
*Note the basic data cleaning and package checking is hidden on the html doc, please see the Rmd file to see how we created the various columns we are using*

```{r echo=FALSE, results='hide', message=FALSE}
if (!require("readxl")) {
install.packages("readxl")
}
if (!require("ggplot2")) {
install.packages("ggplot2")
}
library(readxl)
library(ggplot2)
dat<- readxl::read_xls('GSSlab3data.xls')

```

```{r echo=FALSE, results='hide', message=FALSE}
#the follwing is some data cleaning
dat1 <- dat[!is.na(dat$'Marital status'), ] #removes null values
dat1 <- dat1[dat1$`Age of respondent` != 'Don\'t know',] #following removes age values that dont make sense to plot
dat1 <- dat1[dat1$`Age of respondent` != '89 or older',]
dat1 <- dat1[dat1$`Age of respondent` != 'No answer',]


#the following is creating and manipulating columns
dat1$married <- factor(dat1$`Marital status`, levels= c('Never married', 'Married','Widowed','Divorced','Separated','No answer'))

dat1$happy <- factor(dat1$`General happiness`, levels = c('Very happy','Pretty happy','Not too happy','No answer','Not applicable','Don\'t know'))

dat1$sex <- factor(dat1$`Respondents sex`, levels = c('Female','Male'))

dat1$age <- as.numeric(dat1$`Age of respondent`)

dat1$politicalParty <- factor(dat1$`Political party affiliation`, levels = c('No answer','Don\'t know', 'Other party','Strong republican','Not str republican',
                                                                             'Ind,near rep','Independent','Ind,near dem','Not str democrat','Strong democrat'))

dat1$health <- factor(dat1$`Condition of health`, levels = c('Excellent', 'Good', 'Fair', 'Poor', 'Don\'t know', 'No answer', 'Not applicable'))
```

**Question 1)**
```{r}
ggplot(data = dat1, aes(x=married, fill=happy)) + geom_bar(position = 'Dodge') +xlab("Marital Status") + ylab("Number of respondents") + labs(title="Number of Responents vs Marital Status by Happiness level")
```

Looking at the bar graph we can see a few common patterns in that the most common level of happiness is just pretty happy. Beyond that the ratio of very happy to everything else is higher for married respondents than for the not married respondents and the ratio of not to hapy rises for those who were married but arent anymore. This suggests that married people are happier and that having married and then separating or divorcing leads to an unhappy life.


**Question 2)**
```{r}
ggplot(data = dat1, aes(x=married, fill=happy)) + geom_bar(position = 'Dodge') + theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) + facet_wrap(vars(sex)) +xlab("Marital Status") + ylab("Number of respondents") + labs(title="Number of Responents vs Maritial Status by Happiness level, faceted by sex")
```

Comparing the two the only real diiference is that across the board men are more likely to respond as "not to happy" and not "very happy" so I would say that no the sex of the respondent does not affect the relationships found in question 1.


**Question idea 1 of 4)**
Does Age affect the political affiliation of a person and how does sex affect this relationship?
```{r}
ggplot(data = dat1, aes(x=politicalParty, y=age)) + geom_boxplot() + theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) + facet_wrap(vars(sex)) +xlab("Political Affliation") + ylab("Age") + labs(title="Boxplot of Age by political Affliation, faceted by sex")
```

Looking at the side by side box plots we can see a fairly obvious trend. The younger someone is the more likly they are to be independent. Additionally it is interesting to not that the party itself doesnt matter with regards to age. The only thing that seems to happen is that as people get older they are more likely to indentify as strongly one way or the other. Looking at the differences caused by sex the only real difference we can see is that older men are more likly to be democrate where as older females are more likly to be strong republicans.


**Question idea 2 of 4)**
How does age impact happiness, and how does Health affect this?

```{r}
ggplot(data = dat1, aes(x=happy, y=age)) + geom_boxplot() + theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) + facet_wrap(vars(health)) +xlab("Happiness") + ylab("Age") + labs(title="Boxplot of Age by happiness, faceted by Health")
```

The relationship between and age seems to not actually matter the median age of all happiness levels are about the same there isnt one group that is happier than another. It interesting to  note that as median age increases the health condition decreases. The most interesting thing is that when people respond that they don't know what their health is like then there is a relationship between age and happiness. Older peole tend to be happier when they dont know their health condition. It seems to be that if you want to stay happy when you get old stop worrying about your health. 
