---
title: "Lab 3"
author: "Eric Gleiter; Brandon Xicon"
date: "2/19/2020"
output: html_document
---

```{r}
if (!require("readxl")) {
install.packages("readxl")
}
if (!require("ggplot2")) {
install.packages("ggplot2")
}
dat<- readxl::read_xls('GSSlab3data.xls')

```

Question 1)
```{r}


dat$married <- factor(dat$`Marital status`, levels= c('Never married', 'Married','Widowed','Divorced','No answer','Separated'))
dat$happy <- factor(dat$`General happiness`, levels = c('Very happy','Pretty happy','Not too happy','No answer','Not applicable','Don\'t know'))
dat$sex <- factor(dat$`Respondents sex`, levels = c('Female','Male'))
ggplot(data = dat, aes(x=married, fill=happy)) + geom_bar(position = 'Dodge') +xlab("Marital Status") + ylab("Number of respondents") + labs(title="Number of Responents by Happiness level")
```
Looking at the bar graph we can see a few common patterns in that the most common level of happiness is just pretty happy. Beyond that the ratio of very happy to everything else is higher for married respondents than for the not married respondents and the ratio of not to hapy rises for those who were married but arent anymore. This suggests that married people are happier and that having married and then separating or divorcing leads to an unhappy life.


Question 2)
```{r}
ggplot(data = dat, aes(x=married, fill=happy)) + geom_bar(position = 'Dodge') + facet_wrap(vars(sex))
```

